---
title: "Introduction to hierarchical clustering"
author: Erika Duan
date: "`r Sys.Date()`"
output:
  github_document:
    toc: true
always_allow_html: true
---  


```{r, echo = FALSE, message = FALSE, warning = FALSE}  
#-----load required packages-----  
if (!require("pacman")) install.packages("pacman")
pacman::p_load(here,  
               tidyverse,
               pheatmap,
               dendextend) # for visualising tables   
```


# Introduction  

Hierarchical clustering is a useful unsupervised machine learning approach when you are interested in understanding how similar or disimilar the units in your dataset are which respect to each other.   

Hierarchical clustering is a greedy algorithm and works by:  

1. Calculating the distance between every pair of observations and storing this in a distance matrix. 
2. Setting each observation as an individual cluster.  
3. Merging the closest pair of points into 1 cluster.   
4. Recalculating the distance between every pair of observations (where the new cluster is a new observation). 
5. Repeating this process until clusters merge into a single cluster (i.e. an agglomeration).   

How the distance between clusters is calculated can vary, with `method = "complete` being the default.  


# Creating a test dataset   

Let's imagine that we are interested in whether co-infections between certain bacteria or viruses are common in the human population. We might look through hospital tests and convert those tests into a serious of binary values based on whether a bacteria or virus was tested positive in the same patient.  

**Note:** The code used to create this dataset can be accessed from the `Rmd` file accompanying this tutorial.   

```{r, echo = FALSE, warning = FALSE}
#-----creating a test dataset-----
# this chunk of code has no other purpose except to provide us with an example dataset    
set.seed(123)

#-----create a series of binary responses-----
H3N2_influenza_infection <- rbinom(n = 30, 
                                   size = 1,
                                   prob = 0.32)

H1N1_influenza_infection <- rbinom(n = 30, 
                                   size = 1,
                                   prob = 0.21)

IVB_influenza_infection <- rbinom(n = 30, 
                                  size = 1,
                                  prob = 0.17)

RSV_infection <- rbinom(n = 30, 
                        size = 1,
                        prob = 0.06)

RV_infection <- rbinom(n = 30, 
                       size = 1,
                       prob = 0.11)

adenovirus_infection <- rbinom(n = 30, 
                               size = 1,
                               prob = 0.03)

coronavirus_infection <- rbinom(n = 30, 
                                size = 1,
                                prob = 0.17)

Staphylococcus_aureus_infection <- rbinom(n = 30, 
                                          size = 1,
                                          prob = 0.21)

Streptococcus_pneumoniae_infection <- rbinom(n = 30, 
                                             size = 1,
                                             prob = 0.17)

#-----create final data frame-----
# get() returns the value of a named object
infection_status <- sapply(ls(pattern = "_infection"), get) # converts to matrix
infection_status <- as.data.frame(infection_status)
rownames(infection_status) <- paste0("Pt_", seq(1, 30, 1))
```

```{r}
#-----using kable to quickly visualise the test dataset-----  
infection_status %>%
  head(10) %>%
  knitr::kable()
```


# Perform hierarchical clustering  

```{r}
#-----create distance matrix-----
infection_matrix <- infection_status %>%
  as.matrix() 

infection_dist <- dist(infection_status, method = "euclidean")  
infection_hclust <- hclust(infection_dist, method = "complete")  

infection_hclust$order #shows the order of hierarchical clustering  
```

```{r}
plot(infection_hclust,
     hang = -1) # all labels printed at the bottom  

```

```{r}
# Cut tree into 4 groups
patient_clusters <- cutree(infection_hclust, k = 6)  

# add the cluster that the data belongs to 
infection_status <- infection_status %>%
  mutate(cluster = patient_clusters)
```

https://www.datacamp.com/community/tutorials/hierarchical-clustering-R  

https://www.r-bloggers.com/hierarchical-clustering-in-r-2/ 

https://uc-r.github.io/hc_clustering  

