---
title: "Working with dummy variables and factors"
author: Erika Duan
date: "`r Sys.Date()`"
output:
  github_document:
    toc: true
---

```{r, message = FALSE, warning = FALSE}  
#-----load required packages-----  
if (!require("pacman")) install.packages("pacman")
pacman::p_load(here,
               tidyverse,
               fastDummies) # for creating dummy variables
```


# Introduction   

I found myself wrangling factor levels for a good 30+ minutes earlier this week. The need arose from different machine learning models requiring different data input requirements.   

Let's say you have a binary classification problem and you want to test a random forest model, a gradient boosted tree (XGBoost) model and a support vector model (SVM) model.  

**Requirements:**  

+ The random forest model handles both numerical categorical and numeric variables. The response variable needs to be a factor.     

+ The XGBoost model only handles numerical variables. The predictor variable needs to be converted into a numerical vector (i.e. `0` or `1`).    

+ The linear SVM model requires scaled numerical variables. The predictor variable needs to be converted into a factor that is not represented as `0` or `1`.         

Let's explore a fun test dataset.      


# Creating a test dataset   

Cats are such idiosyncratic creatures. You can never tell if some cats wants a pat or to bite you instead. So what if machine learning could help us better understand cats (to pat or not to pat - that is the question)?     

**Note:** The code used to create this dataset can be accessed from the `Rmd` file accompanying this tutorial.   

```{r, echo = FALSE}
#-----creating a test dataset-----
# this chunk of code has no other purpose except to provide us with an example dataset    

set.seed(111) 

#-----create a vector of random cat breeds-----
cat_breed <- sample(c("british_shorthair",
                      "ragdoll",
                      "bengal",
                      "siamese",
                      "aristocat",
                      "mixed"),
                    size = 500,
                    replace = T)

#-----create a vector for age (rounded to 1 decimal point)-----  
age <- runif(500, 0.5, 9)  
age <- round(age, digits = 1)  

#-----create a vector for corresponding favourite activities-----
fav_activity <- sample(c("window_watching",
                         "hunting_toys",
                         "sitting_on_humans",
                         "napping"),
                       size = 500,
                       prob = c(0.10,
                                0.15,
                                0.25,
                                0.50),
                    replace = T)

#-----create a vector for likes_children-----  
likes_children <- sample(c("yes",
                           "no"),
                         size = 500,
                         prob = c(0.15, 0.85),
                         replace = T)

#-----creating the response variable-----
will_bite <- sample(c("yes", "no"),
                   size = 500,
                   prob = c(0.65, 0.35),
                   replace = T) 

#-----reconstruct two relationship in the dataset-----
cat_prediction <- tibble(cat_breed,
                         age,
                         fav_activity,
                         likes_children,
                         will_bite) 

cat_prediction <- cat_prediction %>%
  mutate(will_bite = case_when(likes_children == "yes" ~ "no",
                               TRUE ~ will_bite),
         will_bite = case_when(cat_breed %in% c("ragdoll", "siamese") ~ "no",
                               TRUE ~ will_bite))
```

```{r}
#-----using kable to quickly visualise the test dataset-----  
cat_prediction %>%
  head(10) %>%
  knitr::kable()
```

# Working with factors


## Creating factors  

There are two ways to handle factors:  

+ Using base `R` to create factors and modify factor levels.    
+ Using [`forcats`](https://r4ds.had.co.nz/factors.html) from the `tidyverse` library to simplify complex factor wrangling.      

Using `factor` without specifying factor levels automatically converts character vectors into a numerical index of character vectors. Specifying factor levels also allows you to specify the order of levels.  

**Note:** Factor levels need to match `unique(vector_values)` to prevent the generation of `NA` values.     

```{r}
#-----converting the response variable into a factor using base R-----  
cat_prediction_factor <- factor(cat_prediction$will_bite)

# str(cat_prediction$will_bite)
#> Factor w/ 2 levels "no","yes": 1 1 2 1 2 2 1 2 2 1 ...  

cat_prediction_factor <- factor(cat_prediction$will_bite,
                                   levels = c("no", "yes")) 

# make sure levels == unique(cat_prediction$will_bite)
# otherwise your response variable will be converted into NAs

cat_prediction_char <- as.character(cat_prediction_factor)  

# str(cat_prediction_char)
#> chr [1:500] "no" "no" "yes" "no" "yes" "yes" "no" "yes" "yes" "no" ...
```

Applying `as.character` back onto a factor returns the original character vector, rather than the numerical index.    

## Modifying factor levels  

In base R, modifying factor levels directly modifies the factor itself. I personally think that the safest way is to modify factor levels is by specifying replacement and original level names using `list`. Note that this approach requires all levels to be specified inside the list, or `NA` values will be generated.       

```{r}
#-----converting factor levels using base R-----
levels(cat_prediction_factor) <- list(no_biting = "no",
                                      yes_ouch = "yes")

# str(cat_prediction_factor)
#> Factor w/ 2 levels "no_biting","yes_ouch": 1 1 2 1 2 2 1 2 2 1 ...

cat_prediction_char <- as.character(cat_prediction_factor)

# str(cat_prediction_char)
#>  chr [1:500] "no_biting" "no_biting" "yes_ouch" "no_biting" "yes_ouch" ...

#----the dangers of modifying levels without referencing names----- 
cat_prediction_factor <- factor(cat_prediction$will_bite,
                                   levels = c("no", "yes")) 
 
# str(cat_prediction_factor)
#> Factor w/ 2 levels "no","yes": 1 1 2 1 2 2 1 2 2 1 ...

levels(cat_prediction_factor) <- c("yes", "no") # accidentally inverting factor level order
 
# str(cat_prediction_factor)
#> Factor w/ 2 levels "yes","no": 1 1 2 1 2 2 1 2 2 1 ...
```

You can also modify factor levels using [`fct_recode`](https://r4ds.had.co.nz/factors.html#modifying-factor-levels
) from `forcats`. The chief advantages of `fct_recode` are that:    

+ It does not modify levels that are not referenced (instead of converting them to `NA`).  
+ It will warn you if you accidentally refer to a level that does not exist.    

```{r}
#-----converting factor levels using forcats-----   
cat_prediction <- cat_prediction %>%
  mutate(will_bite_factor = factor(will_bite),
         will_bite_factor = fct_recode(will_bite,
                                       "no_biting" = "no",
                                       "yes_ouch" = "yes")) # new level name = old level name
```

**Note:** The `forcats` package also contains other functions for modifying or ordering factor levels, which are very useful for plotting tidy graphs with categorical variables. This tutorial is limited to applying factors to binary data.   


# Using dummy variables  

## Creating dummy variables    

According to [Wikipedia](https://en.wikipedia.org/wiki/Dummy_variable_(statistics)),  a dummy variable is one that takes only the value 0 or 1 to indicate the absence or presence of some categorical effect that may be expected to shift the outcome.   

```{r}

```


https://r4ds.had.co.nz/factors.html  
http://www.cookbook-r.com/Manipulating_data/Renaming_levels_of_a_factor/  
