---
title: "Labour force analysis by industry"
output: html_document 
params: 
  industry: "Agriculture, Forestry and Fishing"   
  before_gfc: "2007-05-01"
  before_covid: "2020-02-01"
---

```{r setup, include = FALSE}
if (!require("pacman")) install.packages("pacman")
pacman::p_load(here,  
               tidyverse,
               RColorBrewer)  

# source(here("src", "01_load-and-clean-ABS-data.R"), local = knitr::knit_global())  

tidy_data <- read_csv(here("output", "tidy_ABS_labour_force_by_industry_table_4.csv"))  
```


# `r params$industry` {.tabset}

*Last updated on `r Sys.Date()`.*    

<br>  

```{r, echo = FALSE}
#-----summarise recent changes in employment------  
industry_before_covid <- tidy_data %>%
  filter(industry == params$industry,
         date == params$before_covid) %>%
  pull(count)  

industry_now <- tidy_data %>%
  filter(industry == params$industry,
         date == max(date)) %>%
  pull(count)   

change_since_covid <- 100 * (industry_now / industry_before_covid) - 100 
change_since_covid <- round(change_since_covid , digits = 2)
```

A total of `r format(industry_before_covid * 1000, big.mark = ",")` individuals were employed in `r params$industry` in Feb 2020, before the onset of COVID-19 spread in Australia.        

A total of `r format(industry_now * 1000, big.mark = ",")` individuals were employed in the most recent quarter of `r max(tidy_data$date) %>% format(., "%b %Y")`, which is `r if_else(industry_now > industry_before_covid, "an increase", "a decrease")` of `r if_else(industry_now > industry_before_covid, print(change_since_covid), print(-1 * change_since_covid))` percentage points since Feb 2020.       

<br>  


## Change in employment growth/decline        

<br>  

We can visualise the change in employment for `r params$industry` compared to all other industries as the rate of employment growth or decline since `r min(tidy_data$date) %>% format(., "%b %Y")`.   

<br>  

```{r, echo = FALSE} 
#-----create relative changes plot-----
relative_proportions <- tidy_data %>%
  group_by(industry) %>%
  mutate(count_rate = count/ first(count)) %>%
  ungroup() 

single_industry <- relative_proportions %>%
  filter(industry == params$industry)

getPalette <- colorRampPalette(brewer.pal(9, "Greys"))
total_colours <- length(unique(relative_proportions$industry))

ggplot(relative_proportions, aes(x = date, y = count_rate)) +
  geom_line(aes(colour = industry)) +
  scale_colour_manual(values = getPalette(total_colours)) +  
  geom_line(data = single_industry, col = "firebrick", size = 1.5) + 
  labs(x = NULL,
       y = "Rate of employment growth or decline since Nov 1984",
       title = paste(params$industry)) +
  theme_bw() +
  theme() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.minor.y = element_line(linetype = "dashed"),
        legend.position = "none")
```


## Number of employed persons over time       

<br>  

We can visualise the total number of employed persons for `r params$industry` since `r min(tidy_data$date) %>% format(., "%b %Y")`.      

<br>  

```{r, echo = FALSE}
#-----create total employed persons plot-----  
tidy_data %>%
  filter(industry == params$industry) %>%
  ggplot(aes(x = date, y = count)) +
  geom_line(colour = "firebrick", size = 1) +
  geom_vline(xintercept = c(as.Date(params$before_gfc), as.Date(params$before_covid)),
             colour = "grey50", linetype = "dashed", 
             size = 0.5) +
  labs(x = NULL,
       y = "Total employed persons (1,000s)",
       title = paste(params$industry)) + 
  theme_bw() + 
  theme() + 
  theme(panel.grid.major = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.minor.y = element_line(linetype = "dotted")) +
  annotate("text", x = as.Date("2009-09-01"), y = 440,
           label = "Pre-GFC", colour = "grey50", size = 4.5) +   
  annotate("text", x = as.Date("2016-06-01"), y = 440,
           label = "Pre-COVID-19", colour = "grey50", size = 4.5)   
```


## Changes in total persons employed over time    

<br> 

If we subtract the employed persons count by the preceding quarter's count, we can visualise the total increase or decrease in employment numbers. This allows us to directly compare how much employment numbers have fluctuated over time between different industries.       

<br>  

```{r, echo = FALSE}
#-----create change in employed persons plot-----  
fluctuations <- tidy_data %>%
  group_by(industry) %>%  
  mutate(count_change = count - lag(count, default = first(count))) %>%
  ungroup() 

first_count <- tidy_data %>%
  filter(industry == params$industry,
         date == min(date)) %>%
  pull(count)

single_fluctuation <- tidy_data %>%
  filter(industry == params$industry) %>%
  mutate(count_change = count - lag(count, default = first_count))

ggplot(fluctuations, aes(x = date, y = count_change)) +
  geom_line(aes(colour = industry)) +
  scale_colour_manual(values = getPalette(total_colours)) +  
  geom_line(data = single_fluctuation, col = "firebrick", size = 1.5) + 
  geom_vline(xintercept = c(as.Date(params$before_gfc), as.Date(params$before_covid)),
             colour = "grey50",
             linetype = "dashed") +
  labs(x = NULL,
       y = "Gain or loss in employed persons (1,000s)",  
       title = paste(params$industry)) + 
  theme_bw() + 
  theme() + 
  theme(panel.grid.major = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.minor.y = element_line(linetype = "dashed"),
        legend.position = "none") +
  annotate("text", x = as.Date("2009-09-01"), y = 130,
           label = "Pre-GFC", colour = "grey50", size = 4.5) +   
  annotate("text", x = as.Date("2016-06-01"), y = 130,
           label = "Pre-COVID-19", colour = "grey50", size = 4.5)  
```


# {-}  

**Note**: The trend for `r params$industry` is highlighted in red.   