---
title: "Untitled"
author: "Erika Duan"
date: "21/10/2020"
output: html_document
---

If you are using categorical data you can use the Kruskal-Wallis test (the non-parametric equivalent of the one-way ANOVA) to determine group differences. If the test shows there are differences between the 3 groups. You can use the Mann-Whitney test to do pairwise comparisons as a post hoc or follow up analysis. Since you're only doing a few comparisons (i.e. X vs. Y, X, vs Z, and Y vs. Z) you wouldn't have to worry about family wise error rates.  

With SPSS, you can run a chi-square, and test for pair-wise differences between the pair of groups. Put your 3 groups in columns, and "ride the tube: yes/no in the rows.  Use the raw numbers in the 6 cells.  Select the option to use Bonferroni corrections for the pairwise comparisons. Using sub-scripted (super-scripted?) letters, SPSS will tell you which of the pairwise differences are significant, while preserving the overall (experiment-wise) 0.05 significance level.   

	Chi-square test
	
	Compare three or more unmatched groups	One-way ANOVA	Kruskal-Wallis test	Chi-square test
	
	https://www.graphpad.com/support/faqid/1790/
	
	The chi-square test on the counts actually does check for homogeneity of proportions.  
	
	http://www.sthda.com/english/wiki/two-proportions-z-test-in-r
	
	https://stats.stackexchange.com/questions/70107/fishers-exact-test-in-r-2x4-table 
	
	https://data.library.virginia.edu/pairwise-comparisons-of-proportions/    
	
	https://uc-r.github.io/multivariate_inference  
	
	https://stats.libretexts.org/Bookshelves/Introductory_Statistics/Book%3A_Introductory_Statistics_(Shafer_and_Zhang)/09%3A_Two-Sample_Problems/9.04%3A_Comparison_of_Two_Population_Proportions 
	
	
	
	https://uc-r.github.io/multivariate_inference  
	
	
```{r}
# packages used regularly
library(tidyverse)

# full population data
# A tibble: 4 x 4
## # Groups: Attrition [2]
##   Attrition Gender     n   pct
##   <fctr>    <fctr> <int> <dbl>
## 1 No        Female   501 0.406
## 2 No        Male     732 0.594
## 3 Yes       Female    87 0.367
## 4 Yes       Male     150 0.633  

attrition_by_gender = tibble(Attrition = c("No", "No", "Yes", "Yes"),
                             Gender = c("Female", "Male", "Female", "Male"),
                             n = c(501, 732, 87, 150),
                             proportion = c(0.406, 0.594, 0.367, 0.633))
```
	

```{r}

```
	
	