---
title: "Introduction to Chi-squared and F distributions"  
author: "Erika Duan"
date: "`r Sys.Date()`"
output:
  github_document:
    toc: true
    pandoc_args: --webtex   
---

```{r setup, include=FALSE}
# Set up global environment ----------------------------------------------------
knitr::opts_chunk$set(echo=TRUE, results='hide', fig.show='asis', fig.align='center')  
```


```{r load libraries, message=FALSE, warning=FALSE}  
# Load required packages -------------------------------------------------------  
if (!require("pacman")) install.packages("pacman")
pacman::p_load(here,  
               tidyverse,
               patchwork)   
```


# Chi-squared distribution    

The Chi-squared distribution, with degrees of freedom $k$, describes the distribution of $\displaystyle \sum_{1=1}^{k} Z_k^2$ values, where $Z_k$ represents any value from a standard normal distribution.   

The probability density function of a Chi-squared distribution is described by the equation below:  
$f(x) = \frac{x^{k/2 - 1}e^{-x/2}}{2^{k/2}\Gamma(k/2)}$ for $x \geq 0$.    


## 1 degree of freedom  

The Chi-squared distribution with 1 degree of freedom is simply the distribution of $Z^2$ values.  

```{r plot Z dist, results='markup', fig.width=4, fig.height=3, out.width='50%'}
# Calculate density for values in a standard normal distribution ---------------
Z <- seq(-3, 3, length.out = 1000)
Z_density <- dnorm(Z, mean = 0, sd = 1)

Z_dist <- tibble(values = Z,
                 density = Z_density)

# Plot standard normal distribution --------------------------------------------
Z_dist %>%
  ggplot(aes(x = values, y = density)) +
  geom_line() +
  labs(x = "Z",
       y = "Probability density",
       title = "Standard normal distribution") + 
  theme_minimal() +
  theme(panel.grid = element_blank(),
        panel.border = element_rect(fill = NA, colour = "black"),
        plot.title = element_text(hjust = 0.5))   
```

```{r plot Chisq dist}
# Calculate and plot frequency of Z^2 values -----------------------------------
Z_squared <- Z^2

freq_plot <- Z_squared %>%
  as_tibble() %>%
  ggplot(aes(x = value)) +
  geom_histogram(binwidth = 0.1, fill = "white", colour = "black") +
  labs(x = "Z squared",
       y = "Frequency",
       title = "Frequency of Z squared values") + 
  theme_minimal() +
  theme(panel.grid = element_blank(),
        panel.border = element_rect(fill = NA, colour = "black"),
        plot.title = element_text(hjust = 0.5)) 

# Calculate and plot Chi-squared distribution with df = 1 ----------------------
Chisq <- seq(0, 9, length.out = 1000)

Chisq_density <- dchisq(Chisq, df = 1) 

Chisq_dist <- tibble(values = Chisq,
                     density = Chisq_density)  

density_plot <- Chisq_dist %>%
  ggplot(aes(x = values, y = density)) +
  geom_line() +
  scale_y_continuous(limits = c(0, 1.5)) + 
  labs(x = "Z squared",
       y = "Probability density",
       title = "Chi-squared distribution with df = 1") + 
  theme_minimal() +
  theme(panel.grid = element_blank(),
        panel.border = element_rect(fill = NA, colour = "black"),
        plot.title = element_text(hjust = 0.5))  

(freq_plot + density_plot)
```

**Note:** The shape of this distribution can be explained by the observation that the majority of Z values are distributed around 0 and produce a smaller value when they are squared e.g. $0.1^2 = 0.01$.   


## 2 or more degrees of freedom  


# F distribution     


# Other resources  

+ A jbstatistics [YouTube video](https://www.youtube.com/watch?v=hcDb12fsbBU) on the Chi-squared distribution.  
+ A zedstatistics [YouTube video](https://www.youtube.com/watch?v=80ffqpZdKiA) on the Chi-squared distribution.   
+ The Free University of Berlin [statistics tutorial](https://www.geo.fu-berlin.de/en/v/soga/Basics-of-statistics/Continous-Random-Variables/Chi-Square-Distribution/Chi-Square-Distribution-in-R/index.html) on how to calculate and plot the Chi-squared distribution.   
     
