---
title: "Labour force report for `r paste(params$sex, params$measure)` individuals" 
output: html_document
params:
  sex: "all" 
  measure: "part-time"    
---

```{r, include=FALSE}
# Load required packages -------------------------------------------------------
if (!require("pacman")) install.packages("pacman")
pacman::p_load(here,
               tidyverse,
               clock, 
               gt)  

# Load labour_force_clean dataset ----------------------------------------------
labour_force <- read_csv(here("tutorials",
                              "p-automating_rmd_reports",
                              "abs_labour_force_report",
                              "data",
                              "clean_data",
                              "labour_force_clean.csv"))
```


### Graph of labour force changes since 2019  

This report displays the adjusted monthly change in labour force numbers for `r paste(params$measure, params$sex)` individuals from `r min(labour_force$time_period) %>% str_extract("^[0-9]{4}-[0-9]{2}")` to `r max(labour_force$time_period) %>% str_extract("^[0-9]{4}-[0-9]{2}")`.  

```{r, echo=FALSE}
# Plot data --------------------------------------------------------------------
# Fix y-axis between different reports 
y_max <- max(labour_force$change_obs_value) 
y_min <- min(labour_force$change_obs_value)   

labour_force %>% 
  filter(sex == params$sex, 
         measure == params$measure) %>%
  ggplot(aes(x = time_period, 
             y = change_obs_value)) +
  geom_line() + 
  scale_y_continuous(limits = c(y_min, y_max)) +
  geom_vline(xintercept = as.Date("2020-02-01"),
             colour = "firebrick",
             linetype = "dashed") +
  annotate("label",
           x = as.Date("2020-02-01"),
           y = y_max - 10,
           label = "COVID-19", color = "firebrick") +
  labs(title = paste("Labour force change for", params$sex, params$measure, "individuals"), 
       x = NULL,
       y = "Individuals (1000s)") +
  theme_bw() +
  theme(panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.major.y = element_line(linetype = "dotted"),
        panel.grid.minor.y = element_line(linetype = "dotted"),
        plot.title = element_text(hjust = 0.5))
```


### Table of labour force numbers and changes in the last 12 months  

This table displays the adjusted labour force numbers and adjusted change in labour force numbers for `r paste(params$measure, params$sex)` individuals for the last 12 months, from `r Sys.Date() %>% add_months(-12) %>% str_extract("^[0-9]{4}-[0-9]{2}")` to `r max(labour_force$time_period) %>% str_extract("^[0-9]{4}-[0-9]{2}")`.     

```{r, echo=FALSE}
# Display table ----------------------------------------------------------------
# Create HTML decimal references for triangles
up_arrow <- "<span style=\"color:steelblue\">&#9650;</span>"
down_arrow <- "<span style=\"color:salmon\">&#9660;</span>"

# Create filter to display only the last 6 months of records
six_months_ago <- Sys.Date() %>% add_months(-12)

# Specify table style
labour_force %>% 
  filter(sex == params$sex, 
         measure == params$measure,
         time_period >= six_months_ago) %>%
  arrange(time_period) %>%
  mutate(year =  get_year(time_period),
         month = get_month(time_period)) %>% 
  select(-c(sex, measure, time_period, last_obs_value)) %>% 
  relocate(year, month) %>% 
  rename(number = obs_value,
         change = change_obs_value) %>%
  gt(groupname_col = "year",
     rowname_col = "month",) %>%
   fmt_number(columns = c(number, change),
              decimals = 3) %>% 
    text_transform(locations = cells_body(columns = change, rows = change > 0),
                   fn = function(x) paste(x, up_arrow)) %>%
  text_transform(locations = cells_body(columns = change, rows = change < 0),
                 fn = function(x) paste(x, down_arrow)) %>%  
  tab_header(title = paste(params$sex,
                           params$measure,
                           "individuals") %>% str_to_sentence()) %>%
  tab_footnote(footnote = "Units in 1000s",
               cells_column_labels(columns = number)) %>%
  opt_footnote_marks(marks = "letters") %>%
  tab_footnote(footnote = "Units in 1000s",
               cells_column_labels(columns = change)) %>%
  opt_footnote_marks(marks = "letters") %>%
   tab_options(table.font.size = "small")
```


***  
Report last updated on `r Sys.Date()` from the [ABS labour force dataset](https://www.abs.gov.au/statistics/labour/employment-and-unemployment/labour-force-australia/latest-release)  