---
title: "Fundamental statistical distributions"  
author: "Erika Duan"
date: "`r Sys.Date()`"
output:
  github_document:
    toc: true
    pandoc_args: --webtex   
---

```{r setup, include=FALSE}
# Set up global environment ----------------------------------------------------
knitr::opts_chunk$set(echo=TRUE, results='hide', fig.show='asis')  
```


```{r load libraries, message=FALSE, warning=FALSE}  
# Load required packages -------------------------------------------------------  
if (!require("pacman")) install.packages("pacman")
pacman::p_load(here,  
               tidyverse,
               patchwork)   
```


# Normal distribution    

The normal distribution, or Gaussian distribution, is frequently observed in nature and is characterised by the following properties.  

+ It is a symmetrical continuous distribution where $-\infty < x < \infty$.   
+ Its mean, median and mode are identical due to its symmetrical distribution.  
+ Its probability density function contains two parameters; the mean i.e. $\mu$ and the standard deviation i.e. $\sigma$ where $-\infty < \mu < \infty$ and $\sigma > 0$.   
+ Its probability density function is described by the equation below:   
$f(x) = \frac{1}{\sqrt{2 \pi \sigma^2}} \times e^{\frac{-(x - \mu) ^2}{2\sigma2}}$.    


```{r}
# Calculate expectation of normal distribution ---------------------------------
# Store x * f(x) where f(x) is a normal distribution with mean 140 and sd 16      
funs_x_fx <- function(x) x * (1/ (sqrt(2 * pi * 16 ^ 2)) * exp(-((x - 140) ^ 2) / (2 * 16 ^ 2)))

expectation_n1 <- integrate(funs_x_fx, lower = 92, upper = 188)
expectation_n1  
#> 139.622 with absolute error < 0.00013  
```



## Standard normal distribution  


## Multi-variate normal distribution  



# Chi-squared distribution  


# T distribution   


# F distribution     


# Other resources  

+ University of Sydney Mathematics Learning Centre [chapter](https://www.sydney.edu.au/content/dam/students/documents/mathematics-learning-centre/normal-distribution.pdf) on the normal distribution.     
+ A jbstatistics [YouTube video](https://www.youtube.com/watch?v=iYiOVISWXS4) on the normal distribution.   