---
title: "How to integrate SQL and R - Part 2"  
author: "Erika Duan"
date: "`r Sys.Date()`"
output:
  github_document:
    toc: true
    pandoc_args: --webtex 
---

```{r setup, include=FALSE}
# Set up global environment ----------------------------------------------------
knitr::opts_chunk$set(echo=TRUE, results='hide', fig.show='hold', fig.align='center')  
```

```{r load libraries, message=FALSE, warning=FALSE}  
# Load required packages -------------------------------------------------------  
if (!require("pacman")) install.packages("pacman")
pacman::p_load(here,  
               tidyverse,
               odbc,
               DBI,
               Rcpp)   
```
 
 
# Introduction   

The second part of my tutorial series about SQL to R workflows is mainly focused on breaking down differences between SQL and R syntax. It also contains a rehash of Emily Riederer's excellent [blog post](https://emilyriederer.netlify.app/post/sql-r-flow/) on designing flexible workflows for SQL to R projects. We will be using the SQL database(s) we set up in the [first tutorial](https://github.com/erikaduan/r_tips/blob/master/tutorials/p-sql_to_r_workflows/p-sql_to_r_workflows_part_1.md) for the exercises below.     

```{r create MS SQL connection}
# Create MS SQL connection -----------------------------------------------------
tsql_conn <- DBI::dbConnect(odbc::odbc(),
                            Driver = "SQL Server Native Client 11.0",
                            Server = "localhost",
                            Database = "sandpit",
                            Trusted_Connection = "Yes")
```


# SQL syntax quirks   

In my opinion, the easiest way of learning SQL code is to remember the sequence of how SQL code is executed. In R, because we can chain multiple operations together, the sequence in which R code is written is usually identical to the sequence in which it is executed.  

```{r display execution order, echo=FALSE, results='hold', out.width='80%'} 
knitr::include_graphics("../../figures/p-sql_to_r_workflows-execution_order.svg")
```

In SQL, the sequence in which SQL code is written does not correspond to the sequence in which it is executed, which leads to common errors like the example below.   

```{r, results='hold'}
# Incorrect SQL code -----------------------------------------------------------
# odbc::dbSendQuery(
#   tsql_conn,
#   "
#   SELECT 
#   student_id AS hobbit_id,
#   first_name, 
#   last_name
#   
#   FROM education.student
#   WHERE hobbit_id <> 2
#   
#   -- this query will generate an error as WHERE is executed before SELECT and
#   -- the renaming of student_id to hobbit_id happens after WHERE 
#   "
# ) 

# Correct SQL code -------------------------------------------------------------
odbc::dbSendQuery(
  tsql_conn,
  "
  SELECT
  student_id AS hobbit_id,
  first_name, 
  last_name
  
  FROM education.student
  WHERE student_id <> 2
  
  -- student_id is first filtered via WHERE and lastly renamed to hobbit_id
  "
) %>%
  odbc::dbFetch() %>%
  knitr::kable()
```


## Writing `GROUP BY` SQL queries  

## Writing SQL subqueries  



# Production friendly workflows  


# Other resources  
+ Emily Riederer's excellent [blog post](https://emilyriederer.netlify.app/post/sql-r-flow/) containing ideas for creating flexible SQL <> R project workflows.   
+ Julia Evan's entertaining [programming zine](https://wizardzines.com/zines/sql/) explaining all things SQL. (Paywalled resource)    


    