---
title: "How to integrate SQL and R - Part 2"  
author: "Erika Duan"
date: "`r Sys.Date()`"
output:
  github_document:
    toc: true
    pandoc_args: --webtex 
---

```{r setup, include=FALSE}
# Set up global environment ----------------------------------------------------
knitr::opts_chunk$set(echo=TRUE, results='hide', fig.show='hold', fig.align='center')  
```

```{r load libraries, message=FALSE, warning=FALSE}  
# Load required packages -------------------------------------------------------  
if (!require("pacman")) install.packages("pacman")
pacman::p_load(here,  
               tidyverse,
               odbc,
               DBI,
               Rcpp)   
```
 
 
# Introduction   

The second part of my tutorial series about SQL to R workflows is mainly focused on breaking down differences between SQL and R syntax. It also contains a rehash of Emily Riederer's excellent [blog post](https://emilyriederer.netlify.app/post/sql-r-flow/) on flexible project workflows for integrating SQL with R data analysis.  


# SQL syntax quirks   

```{r display execution order, echo=FALSE, results='hold', out.width='80%'} 
knitr::include_graphics("../../figures/p-sql_to_r_workflows-execution_order.svg")
```





# Production friendly workflows  


# Other resources  
+ Emily Riederer's excellent [blog post](https://emilyriederer.netlify.app/post/sql-r-flow/) containing ideas for creating flexible SQL <> R project workflows.   
+ Julia Evan's entertaining [programming zine](https://wizardzines.com/zines/sql/) explaining all things SQL. (Paywalled resource)    


    